\bf Title here

### 2020-06-08 paper/paper.tex
add DFO in introduction; rewrite introduction + vis diagnostics

\hypertarget{tour}{%
\subsection{Tour}\label{tour}}

Tour provides a way to explore multivariate data interactively via an
established tour path. A tour path is formed by interpolating between
randomly generated plane. Different types of tours are available depends
on the purpose of the exploration i.e.~a grand tour is suitable for
randomly exploring the data from different angles; a guided tour detects
a particular structure in the data. a manual tour allows user to
manually control the projection \citep{cook1997manual}.

\hypertarget{guided-tour}{%
\subsection{Guided tour}\label{guided-tour}}

Guided tour is usually used in conjunction projection pursuit, a method
coined by \citet{friedman1974projection} to detect ``interesting''
low-dimension projection of multivariate data. A projection pursuit
requires the definition of a projection pursuit index function and an
optimisation routine. The projection pursuit index measures the
``interestingness'' of data defined as its departure from normality.
Numerous indices have been proposed in the literature, including
lengendre index \citep{friedman1974projection}, hermite index
\citep{hall1989polynomial}, natural hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}. An optimisation routine is required to
find the projection basis (thus projection) that maximises the
projection pursuit index. The discussion of existing optimisation
procedure will be discussed in the next section.

Guided tour creates visualisation for the projections found by
projection pursuit by constructing a tour path. An illustration modified
from \citep{buja2005computational} entails the geodesic interpolation
between the plane generated by projection pursuit.

\includegraphics[width=7.75in]{/Users/hzha400/Documents/3.PhD/paper-tour-vis/figures/tour_path_new}

\hypertarget{optimisation-methods-in-projection-pursuit-literature}{%
\subsection{Optimisation methods in projection pursuit
literature}\label{optimisation-methods-in-projection-pursuit-literature}}

As \citet{friedman1974projection} said ``\ldots{}, the technique use for
maximizing the (one- and two-dimensional) projection index strongly
influences both the statistical and the computational aspects of the
procedure.'' The quality of the optimisation procedure largely affect
the tour view and thus, the interesting projection one could possibly
observe. An ideal optimisation procedure needs to have following
characteristics:

\begin{itemize}
\item
  \emph{Being able to handle non-differentiable index function}: the
  index function could be noisy and non-differentiable
\item
  \emph{Being able to optimise with constraints}: the projection matrix
  is restricted to an orthonormal matrix.
\item
  \emph{Being able to reveal both local and global maximum}: Although
  the primary interest is to find the global maximum, distinct local
  structures are also of our interest.
\end{itemize}

Below present three existing methods in tour.

\citet{posse1995projection} presented a random search algorithm that
samples new basis in the neighbourhood of the current basis. The
neighbourhood is defined via the radius of the p-dimensional sphere,
\(c\). The new basis is taken as the target basis if it has higher index
value, or the sampling continues. If no basis is found to have higher
index value after a certain number of tries \(n\), the radius \(c\) is
halved. The algorithm stops when the maximum number of iteration is
attained or the radius \(c\) is less than a pre-determined number.
{[}Pursuit package uses this method and it works great! But I don't
think we implement this - although don't think it is too hard to do
it{]}.

@@

\hypertarget{iterative-algorithm-and-its-diagnostics}{%
\section{Iterative algorithm and its
diagnostics}\label{iterative-algorithm-and-its-diagnostics}}

\hypertarget{tour-components}{%
\subsection{Tour components}\label{tour-components}}

Guided tour, along with other types of tour, has been implemented in the
\emph{tourr} package in R, available on the Comprehensive R Archive
Network at \url{https://cran.r-project.org/web/packages/tourr/}
\citep{wickham2011tourrpackage}. A tour includes two major components: a
\emph{generator} that generating the projection basis according to
projection pursuit and an \emph{interpolator} that performing geodesic
interpolation between the projection basis.

The pseudo-code below illustrates the implementation of guided tour in
the tourr package. Given an projection pursuit index function and a
randomly generated projection basis (current basis), the optimisation
procedure produces a target basis inside \texttt{generator()} Both the
current basis and the target basis will be supplied to
\texttt{tour\_path()} to prepare information needed for constructing a
geodesic path. This information is then used to compute a series of
interpolating bases inside the \texttt{tour()} function. All the basis
will be sent to create animation for visualising the tour in the
\texttt{animate()} function.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{animation <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{()\{}

  \CommentTok{# compute projection basis }
\NormalTok{  tour <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{()\{}

    \CommentTok{# construct bases on the tour path}
\NormalTok{    new_geodesic_path <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{()\{}
\NormalTok{      tour_path <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{()\{}

        \CommentTok{# GENERATOR: generate projection basis via projection pursuit}
\NormalTok{        guided_tour <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{()\{}
\NormalTok{          generator <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{()\{}

            \CommentTok{# define projection pursuit index}
            \CommentTok{# generate the target basis from the current basis via optimisation}
\NormalTok{          \}}
\NormalTok{        \}}

        \CommentTok{# prepare geodesic information needed for interpolating along the tour path}
\NormalTok{      \}}
\NormalTok{    \}}

    \CommentTok{# INTERPOLATOR: interpolate between the current and target basis }
    \ControlFlowTok{function}\NormalTok{()\{}
      \CommentTok{# generate interpolating bases on the geodesic path}
\NormalTok{    \}}
\NormalTok{  \}}

  \CommentTok{# animate according to different display methods}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\hypertarget{diagnostics}{%
\subsection{Diagnostics}\label{diagnostics}}

Visualisation has been widely used for exploring and understanding data.
Visualisation presents information in a graphical manner and often
allows people to see information they would otherwise not seen in the
reporting of numerical summarisation. Visual diagnostics can be
real-time or post-run. Real time diagnostic directly uses the data
produced in the algorithm to produce visual representation and thus
doesn't need to store the data. This section focuses on the definition
and production of post-run diagnostics and the next section discusses
real-time diagnostics.

Post-run diagnostics requires the data produced during the algorithm to
be stored/ saved in order to produce plot diagnostics.

The graphical system in R is established based on the conception grammar
of graphic \citep{wickham2010layered}, where a graphic is defined using
stacked layers in a coordination system. A layer includes 1) the dataset
that powers the plot; 2) a geometric object that represents the visual
shape of the plot and 3) relevant statistical transformation if needed.
An important concept in the grammar of graphic is \emph{aesthetic
mapping}. Aesthetic mapping links the variable in a dataset to
information needed to produce a geometric object. For example, we map
one variable on the x-axis and another on the y-axis to create a
scatterplot. To create a boxplot, we first map one variable on the
x-axis and then map the five point summary of another variable on the
y-axis. This computation of five point summary from the origin variable
is statistical transformation. This definition of graphic through layers
provides advantages to produce complex plots since all the plots,
however complex, can be decomposed into basic geometric objects shaped
by the variable supplied.

This idea of decompositing a graph to basic elements inspires me to
characterise the diagnostics of iterative algorithm in a similar
fashion. The optimisation in many machine learning algorithms these days
is iterative in nature, while remains as a black box. Being able to
diagnose it visually allows researchers to have the tool to unfold the
myth and thus provides more understand to the algorithm.

The concept of grammar of graphic requires a dataset to be supplied in
making a plot, thus a global object needs to be created in the iterative
algorithm. The variables will be mapped into element of a graphic to
explore and thus should contain all the parameters of interest. The next
section shows how a global object is created in projection pursuit
guided tour.

\hypertarget{data-structure}{%
\subsection{Data Structure}\label{data-structure}}

In the current implementation of the \texttt{tourr} package, while the
target basis generated by the projection pursuit can be accessed later
via \texttt{save\_history()}, interpolating bases and those randomly
nearby bases generated in the optimisation are not stored. This creates
difficulties for fully understand the behaviour of the optimisation and
interpolation of tour in complex scenario \textbf{{[}need a rephrase
this part{]}.}

Two set of simulated data are used in the demonstration of the
visualisation and diagnostics of the tour optimisation. A small dataset
consists of 1000 randomly simulated observations of five variables
(\texttt{x1}, \texttt{x2}, \texttt{x8}, \texttt{x9}, \texttt{x10}).
\texttt{x2} is the informative variable simulated from two bi-modal
normal distribution centred at -3 and 3 with variance being 1 and the
other four are simulated from \texttt{N(0,\ 1)}. The data has been
scaled to ensure \texttt{x2} has variance of 1.

A larger dataset contains more informative variables (\texttt{x3} to
\texttt{x7}) of different types. \texttt{x3} takes 500 positive one and
500 negative one. The distribution of all the variables except
\texttt{x3} is plotted below. \emph{{[}should I introduce the dist for
each var?{]}}

@@

\hypertarget{post-run-static-diagnostics-plots}{%
\section{Post-run static diagnostics
plots}\label{post-run-static-diagnostics-plots}}

Given all the information of interest in a global object, the next thing
is to design its visual representation that can diagnose our algorithm.
In an iterative algorithm, one would be interested to know how one
parameter changes as the algorithm iterates. The parameter could be
could be the decision variable, the value of the optimisation function
or other parameter that facilitate the algorithm. In projection pursuit
guided tour, apart from the decision variable projection basis, the
value of index function (\texttt{index\_val}), we are also interested to
explore the effect the different searching methods, the neighbourhood
parameter alpha, and the stage a particular observation is in
(\texttt{info}). These are the variables that will be mapped into the
diagnostic plot. The global object also has its time series nature since
the observations are recorded as the algorithm progresses and thus can't
be switch in rows. This time series structure is always hierarchical and
nested since each row can be labelled with the smallest unit \texttt{id}
and up by the iterative structure of the algorithm. Take our example, in
the global object, each row has an \texttt{id} label which is the
smallest unit the observation is ordered. A larger ordering unit is
\texttt{tries} since it is updated every time a new target basis is
found. A nested ordering unit is \texttt{loop}, which increases by one
as searching method iterates and starts over at a new \texttt{tries}.
These features explain the definition of the global object for
projection pursuit guided tour in Figure \ref{fig:glb-obj}.

\begin{figure}
\includegraphics[width=1\linewidth,height=0.2\textheight]{/Users/hzha400/Documents/3.PhD/paper-tour-vis/figures/global_obj} \caption{\label{glb-obj}this is xxxx}\label{fig:glb-obj}
\end{figure}

@@

\emph{Example: exploring searching points} The data is sourced from a 2D
projection of the larger dataset and \texttt{search\_better} is used
with \texttt{max.tries\ =\ 500}. In Figure \ref{fig:points-id} and
\ref{fig:points-tries}, a comparison has been made on visualising the
searching points with \texttt{id} and \texttt{tries} on the x-axis,
colored by \texttt{tries}. In Figure \ref{fig:points-id}, the searching
points of the first few \texttt{tries} are squeezed in a small width
leaving the large uninteresting searching points in the last three
\texttt{tries} taking a vast majority of the plotting space. While in
Figure \ref{fig:points-tries} the data is spaced by \texttt{tries}
evenly in the plot. Label at the bottom indicates the number of
observations in each tries and facilitates the choice of cutoff to
switch from point geometry to boxplot geometry (\texttt{cutoff\ =\ 15}).
The line geometry suggest the largest improvement happens at
\texttt{tries\ =\ 5}.

@@

@@

Matrix parameter could also be interested to explore, for example, in
the tour, we are interested to explore the relative position of the
projection basis in the vector space. A projection basis is a
vector/matrix whose row number is the number of variable in the given
dataset. This imposes difficulties in visualisation since we are bounded
to perceive at most three dimensions. Thus, principal component analysis
is used to reduce the dimension of the projection basis and the first
two principal component are mapped to the x and y axis of the plot.
Another variable of interest could be mapped to the color aesthetics to
see how it changes as the projection basis changes in the projected 2D
space. {[}have a second read here{]}

>>>>

\hypertarget{derivative-free-optimisation}{%
\subsection{Derivative free
optimisation}\label{derivative-free-optimisation}}

Given an objective function \(f\), one way of optimising it is to equate
its gradient to zero. In modern optimisation problems, gradient
information can be hard to evaluate or sometimes even impossible and
Derivative-Free Optimisation (DFO) methods can be useful to approach
these problems. The two most common methods in DFO are
\emph{direct-search methods} and \emph{model-based methods} and this
paper dedicates to the discussion of direct-search methods, which gains
its popularity due to its simplicity in use and reliability in
complicated practical problem. Coined by \citet{hooke1961direct}, direct
search methods don't require any gradient or hessian information and
evaluate \(f\) directly. A well-known example of it is the Nelder-Mead
algorithm \citep{nelder1965simplex} and it enjoys the popularity due to
its simplicity and reliability. {[}feel like this sentence can be
expanded to include more information. xxx{]}. Further, direct search
methods can be classified as \emph{stochastic} and \emph{deterministic}
depends on whether a random sample is used in the search. The stochastic
version of direct-search method is the focus of this paper.

\hypertarget{projection-pursuit-guided-tour}{%
\subsection{Projection pursuit guided
tour}\label{projection-pursuit-guided-tour}}

The optimisation problem we're interested in is in the context of
projection pursuit. Coined by \citet{friedman1974projection}, projection
pursuit is a method that detects the interesting structure
(i.e.~clustering, outliers and skewness) of multivariate data via
projecting it in lower dimensions. A Projection Pursuit Index (PPI) is a
function of the projection matrix (a.k.a projection angle, projection
basis) measuring the ``interestingness'' of data and we refer the value
of PPI function as \emph{index value}. In the literature, the indices
being proposed include lengendre index \citep{friedman1974projection},
hermite index \citep{hall1989polynomial}, natural hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}. {[}Any literature on holes index?
xxx{]}

As \citet{friedman1974projection} noted ``\ldots{}, the technique use
for maximising the projection index strongly influences both the
statistical and the computational aspects of the procedure.'' A suitable
optimisation procedure is needed to find the projection angle that
maximises the PPI and the quality of the optimisation largely affect the
interesting projections one could possibly observe.

Projection pursuit is usually used in conjunction with a tour method
called \emph{guided tour}. Tour explores the multivariate data
\emph{interactively} via playing a series of projections, that form a
\emph{tour path}, like movie frames in real time. Starting from a
randomly generated projection basis, projection pursuit searches for a
target basis with higher index value and guided tour interpolates
geodesically between the initial basis and the target basis. This series
of bases forms a tour path and the projection of the data on these bases
is played in real time. This search-and-interpolateprocess is repeated
until no basis can be found with higher index value. Modified from
\citet{buja2005computational}, Figure \ref{tour-path} vividly depicts
the tour path in guided tour. Guided tour, along with other types of
tour, has been implemented in the \emph{tourr} package in R, available
on the Comprehensive R Archive Network at
\url{https://cran.r-project.org/web/packages/tourr/}
\citep{wickham2011tourrpackage}.

\begin{figure}
\includegraphics[width=1\linewidth,height=0.6\textheight]{/Users/hzha400/Documents/3.PhD/paper-tour-vis/figures/tour_path_increasing} \caption{\label{tour-path}An illustration of the tour path}\label{fig:tour-path}
\end{figure}

{[}doubt if I need the following. xxx{]}

\hypertarget{optimisation-in-projection-pursuit}{%
\subsection{Optimisation in projection
pursuit}\label{optimisation-in-projection-pursuit}}

Below listed several features characterise the optimisation procedures
needed in projection pursuit

\begin{itemize}
\item
  \emph{Being able to handle non-differentiable PPI function}: The PPI
  function could be non-differentiable, thus derivative free methods are
  preferred.
\item
  \emph{Being able to optimise with constraints}: The constraint comes
  from projection matrix being an orthonormal matrix.
\item
  \emph{Being able to find both local and global maximum}: Although the
  primary interest is to find the global maximum, local maximum could
  also reeal structures in the data that are of our interest.
\end{itemize}

{[}thinking if the following three should be presented as algorithms or
plain description is fine. xxx{]}

There are three existing methods for optimisating PPI function and we
review them below. \citet{posse1995projection} used a stochastic direct
search method, a random search algorithm to sample new bases in the
neighbourhood of the current basis defined by the radius of the
p-dimensional sphere, \(c\). The new basis is taken as the target basis
if it has higher index value, or the sampling continues. If no basis is
found to have higher index value after a certain number of tries \(n\),
the radius \(c\) is halved. The algorithm stops when the maximum number
of iteration is attained or the radius \(c\) is less than a
pre-determined number.

@@

\hypertarget{visual-diagnostic-system}{%
\section{Visual diagnostic system}\label{visual-diagnostic-system}}

\hypertarget{visual-diagnostics}{%
\subsection{Visual diagnostics}\label{visual-diagnostics}}

Random search methods has a black-box machinism and focuses solely on
finding the global maximum point, while the projection pursuit problem
we have is aiming at \emph{exploring} the data and thus is interested in
how the algorithm finds its maximum. This motivates us \textbf{to
develop a visual diagnostic system for exploring the optimisation
searching path}.

The necessity of developing such a system rather than simply producing
different diagnostic plots is because there is always needs to develop a
function specific for diagnosing one particular variables and these
functions developed are not reusable at all. Thus we want to establish a
set of rules that prescribe the diagnostic of iterative algorithms
systematically.

The idea of unifying all the diagnostic plots under one framework is
inspired by the concept of \emph{gramma of
graphic}\citep{wickham2010layered}, which powers the primary graphical
system in R, ggplot2 \citep{ggplot2}. In gramma of graphic, plots are
not defined by its appearance (i.e.~boxplot, histogram, scatter plot
etc) but by ``stacked layers''. By this design, ggplot doesn't need to
develop a gazillion of functions that each produces a different type of
plot, instead, it defines layers and allow users to specify the plot
through these defined layers.

{[}Do I need this paragraph to details the gramma of graphic? xxx{]}

\hypertarget{glocal-object}{%
\subsection{Glocal object}\label{glocal-object}}

Ggplot prescribes how to make a graphic but it doesn't prescribe how to
run a diagnostic. {[}need to have a paragraph or two here to describe
how I come up with the design of four components in the global object -
need to related to the component in optimisation. xxx{]}.

A global object used for diagnosing an optimisation is defined with four
components: decision variable, value of objective function, iterators
and other parameter of interest. In the projection pursuit guided tour,
the decision variable is the projection matrix, the value of objective
function is index value. There are three iterators: \texttt{id} is the
smallest ordering unit that increases by one for each observatoin;
\texttt{tries} is a larger order unit, which is updated once a
search-and-interpolate step is finished. A third iterator \texttt{loop}
is used to record the number of repitition in the search step and starts
over from one at the beginning of a new \texttt{tries}. Since there are
different methods used in the projection pursuit guided tour, we are
interested to compare between different methods, thus \texttt{method} is
created as another parameter of interest. The neighbourhood parameter
alpha is always interested in the direct search methods since it
controls the size of the sampling space, thus \texttt{alpha} is another
parameter of interest. \texttt{info} is created to label different stage
in the searching process and it is the third parameter of interest. A
sketch of the global object for projection pursuit guided tour is
presented in Figure \ref{fig:glb-obj}.

\begin{figure}
\includegraphics[width=1\linewidth,height=0.2\textheight]{/Users/hzha400/Documents/3.PhD/paper-tour-vis/figures/global_obj} \caption{\label{glb-obj}The global object in projection pursuit guided tour.}\label{fig:glb-obj}
\end{figure}

\hypertarget{diagnostics-plots}{%
\section{Diagnostics plots}\label{diagnostics-plots}}

\hypertarget{simulated-data}{%
\subsection{Simulated data}\label{simulated-data}}

Two set of simulated data are used in the demo of the visual diagnostics
in projection pursuit guided tour. A small dataset consists of 1000
randomly simulated observations of five variables (\texttt{x1},
\texttt{x2}, \texttt{x8}, \texttt{x9}, \texttt{x10}). \texttt{x2} is the
informative variable simulated from two bi-modal normal distribution
centred at -3 and 3 with variance of one and the other four are
simulated from non-informative standard random normal distributions. The
data has been scaled to ensure \texttt{x2} has variance of 1. The goal
is to find the 1D projection of bi-modal shape, which correspondes to
the projection matrix (vector) of \texttt{{[}0,\ 1,\ 0,\ 0,\ 0{]}}.

A larger dataset contains more informative variables (\texttt{x3} to
\texttt{x7}) of different types. The distribution of all the variables
except \texttt{x3} is plotted below and \texttt{x3} takes 500 positive
one and 500 negative one. {[}should I introduce the dist for each var?
xxx{]} {[}also feel like we don't really use \texttt{x3} to \texttt{x6},
should we not mention about these? xxx{]}

@@

@@

@@

\emph{Example: exploring searching points} The data is sourced from a 2D
projection of the larger dataset and \texttt{search\_better} is used
with \texttt{max.tries\ =\ 500}. In Figure \ref{fig:points-id} and
\ref{fig:points}, a comparison has been made on visualising the
searching points with \texttt{id} and \texttt{tries} on the x-axis,
colored by \texttt{tries}. In Figure \ref{fig:points-id}, the searching
points of the first few \texttt{tries} are squeezed in a small width
leaving the large uninteresting searching points in the last three
\texttt{tries} taking a vast majority of the plotting space. While in
Figure \ref{fig:points} the data is spaced by \texttt{tries} evenly in
the plot. Label at the bottom indicates the number of observations in
each tries and facilitates the choice of cutoff to switch from point
geometry to boxplot geometry (\texttt{cutoff\ =\ 15}). The line geometry
suggest the largest improvement happens at \texttt{tries\ =\ 5}.

@@

Sometimes, a parameter to explore could be a vector or matrix instead of
scalar, for example, in tour, we are interested to explore the
projection basis in the vector space. This imposes difficulties in
visualisation since we are bounded to perceive at most three dimensions.
Thus, principal component analysis is used to reduce the dimension of
projection bases and the first two principal components are mapped to
the x and y axis of the plot. Additional variable of interest could be
mapped to the color aesthetics to for exploration.
